---
title: 'Heatmap & Levels of Granularity'
author: "David Effendi"
date: "Mar 16, 2018"
output: "ioslides_presentation"

---
## Right level of granulairty?

The three heatmaps that follows are built from the same dataset, but at different levels of granularity. 

- Too granular, the plot might be too noisy, pattern is hardly visible
- Too aggregated, might miss some details of the pattern

For each plot, R uses the base dataset, re-group the rows according to the granularity level, aggregating count of incidents for each time interval, and then finally pass it to plotly for plotting.

 Data for this exercise was obtained from [NYPD's open data on Vehicle Motor Collisions](https://data.cityofnewyork.us/Public-Safety/NYPD-Motor-Vehicle-Collisions/h9gi-nx95).

---

### Very Granular (1 minute interval)

```{r, echo=FALSE, results="as-is", warning=FALSE, message=FALSE, fig.height=4}
library( plotly )
### Load the data
inData <- read.table( './NYPD_Motor_Vehicle_Collisions.csv',
                      skip = 1,
                      header = FALSE,
                      stringsAsFacto = FALSE,
                      fill = TRUE,
                      quote = '"',
                      sep = ","
                    ) 
colNames <- c( "DAY",
              "TIME",
              "BOROUGH",
              "ZIP",
              "LATITUDE",
              "LONGITUDE",
              "LOCATION",
              "ON.ST.NAME",
              "CROSS.ST.NAME",
              "OFF.ST.NAME",
              "PERSONS.INJ",
              "PERSONS.KIL",
              "PEDESTRIANS.INJ",
              "PEDESTRIANS.KIL",
              "CYCLISTS.INJ",
              "CYCLISTS.KIL",
              "MOTORISTS.INJ",
              "MOTORISTS.KIL",
              "FACTOR1",
              "FACTOR2",
              "FACTOR3",
              "FACTOR4",
              "FACTOR5",
              "UKEY",
              "VEHICLE1",
              "VEHICLE2",
              "VEHICLE3",
              "VEHICLE4",
              "VEHICLE5"
             )
names(inData) <- colNames

### Discard rows without Contributing Factors
restateNA <- function( x ) {
    ### replace spaces, "unspecified" and blanks with NA
    x <- sub( "^(\\s*|unspecified|)$", NA, x, ignore.case = TRUE )
}

inData$FACTOR1 <- restateNA( inData$FACTOR1 )
inData$FACTOR2 <- restateNA( inData$FACTOR2 )
inData$FACTOR3 <- restateNA( inData$FACTOR3 )
inData$FACTOR4 <- restateNA( inData$FACTOR4 )
inData$FACTOR5 <- restateNA( inData$FACTOR5 )

sel <- which( is.na( inData$FACTOR1 ) &
              is.na( inData$FACTOR2 ) &
              is.na( inData$FACTOR3 ) &
              is.na( inData$FACTOR4 ) &
              is.na( inData$FACTOR5 )           
            )
inData <- inData[-sel,]

### Recode DATE to WEEKDAY
library( lubridate )
inData$WEEKDAYS <-weekdays( mdy(inData$DAY) )
inData$WEEKDAYS <- as.factor(inData$WEEKDAYS)

### Recode TIME to "DAY"(07:00 - 18:59) /"NIGHT (19:00 - 06:59)"
minsSinceMidnight <- function( x ) {
    as.numeric( difftime( as.POSIXct( x, format = '%H:%M'),
                          as.POSIXct('00:00', format = '%H:%M'),
                          units = 'min'
                        )
              )
}
inData$TIME.MINS <- minsSinceMidnight( inData$TIME )

timeGrouper <- function( x, timeTable ) {
### convertes time (mins since midnight) to one of many groups defined in the
### timeTable argument
### format is:   START   END   LABEL
    y <- c()
    for( t in 1:dim(timeTable)[1] ){
        if( x >= timeTable[t,1] && x <= timeTable[t,2] ){ y <- timeTable[t,3] }
    }
    return( y )
}
timeGroups <- data.frame(c(420, 1140, 0000 ), c(1139, 1439, 419 ),c("Day", "Night", "Night"))
inData$TIME.GRP <- sapply( inData$TIME.MINS, FUN=timeGrouper, timeTable=timeGroups )
inData$TIME.GRP <- as.factor(inData$TIME.GRP)

### Recode ISWEEKEND
inData$ISWEEKEND <- ( inData$WEEKDAYS == "Saturday" |
                      inData$WEEKDAYS == "Sunday"   |
                      ( inData$WEEKDAYS == "Friday" & inData$TIME.MINS >= 1140 )
                    )
inData$ISWEEKEND <- as.factor(inData$ISWEEKEND)

### Expand Contributing Factors into long format
factorList <- unique( c(unique(inData$FACTOR1),
                       unique(inData$FACTOR2),
                       unique(inData$FACTOR3),
                       unique(inData$FACTOR4),
                       unique(inData$FACTOR5)
                      )
                   )
factorList <- factorList[ -(which( is.na(factorList) )) ]

addFactor <- function( daColName, daDF, daLookups ) {
    f <- rep( 0, dim(daDF)[1] )
    for(j in 1:length(daLookups) ){
        f <- f + ( daDF[ , daLookups[ j ] ] %in% daColName )
    }
    f[ f >= 1 ] <- 1
    f <- as.factor( f )
    return( f )
}

for(i in 1:length(factorList)){
    nn <- addFactor( factorList[i], inData, c(19:23) )
    cmdd <- paste0("inData$","'", factorList[i] ,"' <- nn")
    eval(parse(text=cmdd))
}

### Recode outcome var from number of persons harmed
inData$HARMFUL <- inData$PERSONS.INJ + inData$PERSONS.KIL
inData$HARMFUL[ inData$HARMFUL>=1 ] <- 1
inData$HARMFUL <- as.factor(inData$HARMFUL)

### save inData to csv
write.csv( inData, "./inData.csv" )

### For HEATMAP of day x time:
library(dplyr)

minsToClock <- function( x ){
### takes a number and converts it to string clock
### e.g. 182 into "03:02"
    res <- NA
    if( x > 0 || x < 1440 ){
        res <- paste0( sprintf( "%02d", x %/% 60 ), ":", sprintf( "%02d", x %% 60 ) )
    }
    return( res )
}

makeYLabels <- function( ival ) {
### returns a vector with size 1440 / ival
### labelled minutes
    ret <- minsToClock( seq( 0, 1439, by=ival ) )
    return( ret )
}

### narrow down to only columns of interest
dat <- inData[,c( 1,2, 30:81 )]
names(dat) <- gsub("[\\/\\s\\(\\)\\_\\-\\']", ".", names(dat), perl=TRUE, )

### group and summarize
datHeatMap <- dat %>% group_by( WEEKDAYS, TIME ) %>%
    summarize( COUNT=n() )


createMissingRows <- function( x, daWeekday ) {
    allMins <- 0:1439
    missingMins <- !(allMins %in% x)
    newMins <- allMins[missingMins]
    nn <- length( newMins )
    newRows <- data.frame(WEEKDAYS=rep( daWeekday, nn),
                          TIME = minsToClock( newMins ),
                          COUNT = rep(0, nn),
                          TIME.MINS = newMins
                         )
    newRows$WEEKDAYS <- as.factor(newRows$WEEKDAYS)
    newRows$TIME <- as.character(newRows$TIME)
    newRows$COUNT <- as.integer(newRows$COUNT)
    newRows$TIME.MINS <- as.numeric(newRows$TIME.MINS)
    return( newRows )
}

addMissingMins <- function( daDF, daWeekday ) {
    newRows <- createMissingRows( subset(daDF, daDF$WEEKDAYS == daWeekday)$TIME.MINS, daWeekday )
    for(i in 1:nrow(newRows)){
        aRow <- as.list(newRows[i,])
        daDF <- daDF %>% rbind( aRow )
    }
    return(daDF)
}
### ensure each weekdays have complete range of minutes
datHeatMap$TIME.MINS <- minsSinceMidnight( datHeatMap$TIME )
datHeatMap <- addMissingMins(datHeatMap, "Monday")
datHeatMap <- addMissingMins(datHeatMap, "Tuesday")
datHeatMap <- addMissingMins(datHeatMap, "Wednesday")
datHeatMap <- addMissingMins(datHeatMap, "Thursday")
datHeatMap <- addMissingMins(datHeatMap, "Friday")
datHeatMap <- addMissingMins(datHeatMap, "Saturday")
datHeatMap <- addMissingMins(datHeatMap, "Sunday")

### recode WEEKDAYS from words into numbers
datHeatMap[ datHeatMap$WEEKDAYS=="Monday",1 ] <- 1
datHeatMap[ datHeatMap$WEEKDAYS=="Tuesday",1 ] <- 2
datHeatMap[ datHeatMap$WEEKDAYS=="Wednesday",1 ] <- 3
datHeatMap[ datHeatMap$WEEKDAYS=="Thursday",1 ] <- 4
datHeatMap[ datHeatMap$WEEKDAYS=="Friday",1 ] <- 5
datHeatMap[ datHeatMap$WEEKDAYS=="Saturday",1 ] <- 6
datHeatMap[ datHeatMap$WEEKDAYS=="Sunday",1 ] <- 7

### sort by day and timeTable
datHeatMap <- datHeatMap %>% arrange( WEEKDAYS, TIME )
```

```{r, echo=FALSE, message = FALSE, warning = FALSE, results="as-is", fig.height=4}
### 1 mins interval (HOURLY)
intvl = 1
datHeatMap$TIME.SEG <- datHeatMap$TIME.MINS %/% intvl
datHeatMapSeg <- datHeatMap %>% group_by( WEEKDAYS, TIME.SEG ) %>%
    summarize( COUNT = sum(COUNT) )%>%
    arrange( WEEKDAYS, TIME.SEG )

dm2 <- matrix( datHeatMapSeg$COUNT, nrow=(1440/intvl), ncol=7 )

p <- plot_ly(
    x = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'),
    y = makeYLabels(intvl),
    z = ~dm2,
    type = "heatmap",
    text = c('Apples', 'Pears', 'Peaches', 'Bananas', 'Pineapples', 'Cherries')
  ) %>%
  layout(
    title = 'Number of incidents by day and time (01/01/2013 to 31/12/2017)',
    xaxis = list(
      type = 'category',
      title = 'Day'
    ),
    yaxis = list(
        title = 'Hour',
        type='bar'
    )
)
p
```
Pattern hardly visible when plotted on a per-minute basis.
---

## Very Aggregated (3 hours interval)
```{r, echo=FALSE, message = FALSE, warning = FALSE, results="as-is", fig.height=4}
### 180 mins interval (HOURLY)
intvl = 180
datHeatMap$TIME.SEG <- datHeatMap$TIME.MINS %/% intvl
datHeatMapSeg <- datHeatMap %>% group_by( WEEKDAYS, TIME.SEG ) %>%
    summarize( COUNT = sum(COUNT) )%>%
    arrange( WEEKDAYS, TIME.SEG )

dm2 <- matrix( datHeatMapSeg$COUNT, nrow=(1440/intvl), ncol=7 )

p <- plot_ly(
    x = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'),
    y = makeYLabels(intvl),
    z = ~dm2,
    type = "heatmap",
    text = c('Apples', 'Pears', 'Peaches', 'Bananas', 'Pineapples', 'Cherries')
  ) %>%
  layout(
    title = 'Number of incidents by day and time (01/01/2013 to 31/12/2017)',
    xaxis = list(
      type = 'category',
      title = 'Day'
    ),
    yaxis = list(
        title = 'Hour',
        type='bar'
    )
)
p
```
Pattern visible ('quiet'/'busy' periods distinguishable, 'Mon-Fri'/'Sat-Sun' differs), but some details may not be reflected
---

## Balanced Level of Granulairty (hourly)

```{r, echo=FALSE, message = FALSE, warning = FALSE, results="as-is", fig.height=4}
### 60 mins interval (HOURLY)
intvl = 60
datHeatMap$TIME.SEG <- datHeatMap$TIME.MINS %/% intvl
datHeatMapSeg <- datHeatMap %>% group_by( WEEKDAYS, TIME.SEG ) %>%
    summarize( COUNT = sum(COUNT) )%>%
    arrange( WEEKDAYS, TIME.SEG )

dm2 <- matrix( datHeatMapSeg$COUNT, nrow=(1440/intvl), ncol=7 )

p <- plot_ly(
    x = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'),
    y = makeYLabels(intvl),
    z = ~dm2,
    type = "heatmap",
    text = c('Apples', 'Pears', 'Peaches', 'Bananas', 'Pineapples', 'Cherries')
  ) %>%
  layout(
    title = 'Number of incidents by day and time (01/01/2013 to 31/12/2017)',
    xaxis = list(
      type = 'category',
      title = 'Day'
    ),
    yaxis = list(
        title = 'Hour',
        type='bar'
    )
)
p
```
Additional detail is now visible, there are two 'peaks' on Mon-Fri, 08:00-09:00 and 16:00-17:00